### 2016-12-06 Projects/DelCo_2012/Report/DelCo_2012.tex
Dan final revisions

\begin{tabbing}
\hspace{1.5in} \= \\
BDP \> Bi-Directional Probe \\
CAF \> Compressed Air Foam \\
CAFS \> Compressed Air Foam System \\
Cal Poly \> California Polytechnic State University, San Luis Obispo \\
DHS \> Department of Homeland Security \\
FEMA \> Federal Emergency Management Administration \\
HF \> Heat Flux \\
IDLH \> Immediate Danger to Life and Health \\
ESTC \> Delaware County Emergency Services Training Center \\
MCFRS \> Montgomery County (MD) Fire and Rescue Service \\
NIST \> National Institute of Standards and Technology \\
OSB \> Oriented Strand Board \\
TC \> Thermocouple \\
TFT \> Task Force Tips \\
TJI \> Trust Joist I-Joist \\
UL FSRI \> UL Firefighter Safety Research Institute \\

@@

Full scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental arrangements and results are presented in this report. The results from both the gas cooling and the suppression of residential scale room fires showed the effectiveness of water and CAF hose streams to be similar. Under the test conditions, both agents were effective. This study did not address the previously documented attributes of CAF for improved exposure protection and improved prevention of a fire re-kindle relative to plain water.

@@

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A aqueous solution into an aerated form (a blanket of bubbles) that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they had applied the agent so that they could make a more effective application. In the wildland, water for fire fighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire or can be used for an exterior (indirect) fire suppression attack on a structure.

@@

\begin{table}[!ht]
\centering
\captionof{table}{Tests to Assess CAF Fire Fighting Agent Compared to Water}\label{tab:agent_tests}
\begin{tabular}{lcc}
\toprule[1.5pt]
Test                                    & Similar Performance & Enhanced Performance \\
\midrule
Specific Heat                           &  \checkmark         &                      \\
Fuel Cooling and Penetration            &                     &  \checkmark          \\
Mass Retention                          &                     &  \checkmark          \\
Ignition Inhibition                     &                     &  \checkmark          \\
Drop Size                               &  \checkmark         &                      \\
Contact Angle                           &                     &  \checkmark          \\
Tire Fire Suppression                   &  \checkmark         &                      \\
Wood Crib Suppression                   &  \checkmark         &                      \\
Heptane Fire Suppression                &  \checkmark         &                      \\
Magnesium \& Titanium Fire Suppression  &  \checkmark         &                      \\
Smoke Generation                        &  \checkmark         &                      \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland fire fighting, CAFS is typically applied directly to the fuel from a remote location, typically an environment that is not an immediate danger to life and health (non-IDLH environment). In structural fire fighting, crews typically have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior fire fighting with CAF are better understood.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires while CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of MCFRS issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order was a catalyst for this study.

@@

The fire suppression series included 20 experiments. Twelve experiments were conducted in the single-story structures (6 in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied and included straight stream, solid stream, and narrow fog stream. The six experiments in the two-story structure used both water and CAF from a straight stream nozzle.  In addition, two CAF experiments were conducted in the two-story burn building with a solid stream.

@@

Each structure had a burn room, approximately 6.0~m (19.6~ft) by 4.4~m (14.3~ft) and 2.74~m (9.0~ft) in height and a hallway connecting the burn room to the front of the structure via a small entry foyer. The hallway was 0.95~m (3.1~ft) wide and 5.4~m (17.6~ft) long and the entry foyer was 1.2~m (4.1~ft) by 1.9~m (6.1~ft). The ceiling height in the hallway and entry foyer was 2.4~m (8~ft).  The open door on north side from the exterior to the entry foyer was 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide. The opening on the south face of the structure from the exterior to the burn room was 2.4~m (7.75~ft) high and 1.09~m (3.6~ft) wide. A basic plan view of the structure is presented in Figure~\ref{fig:Test_Structure_Floor_Plan}, and a fully dimensioned plan view of the structure is included in in Appendix~\ref{app:floor_plans}.

@@

A second story was added to the west single-story structure in order to test larger fires and remote suppression effects (see Figure~\ref{fig:delco_2story}).

@@

The fuel load for test numbers 1 through 8 consisted of two sleeper sofas, two cushioned chairs, carpet, padding, and wood paneling along the southeast corner walls. An image of the fuel arrangement is shown in Figure~\ref{fig:Furniture_Fuel_Load}. The couches and chairs were set up to mimic a typical seating area with one couch against the south wall and one couch against the east wall, each positioned 0.9~m (36~in) away from the southeast corner. A chair was placed adjacent to each couch. The specific locations can be found in the dimensioned schematic, Figure~\ref{fig:Furniture_Fuel_Load_Dimensions}. In total, this fuel package contains approximately 336~kg (739~lb) of fuel consisting of wood, flexible polyurethane (PU), polyester (PE),and polypropylene (PP). Table~\ref{tab:Fire_Suppression_Fuel_Masses} provides a breakdown on the fuel composition for these experiments.

@@

The fuel load for experiments 13 and 14 consisted of three sofas with an average mass of 48.7~kg; 51.4~m$^2$ (559~ft$^2$) of carpet and padding; and approximately 38 sheets of OSB paneling with dimensions 16~mm (0.63~in) x 1.2~m (4~ft) x 2.4~m (8~ft) along the east, west, and south walls and ceiling. The three couches were aligned along the west wall and spaced 1.2~m (4~ft) apart. The specific locations can be found in the dimensioned schematic Figure~\ref{fig:furniture_2story}. In total, this fuel package was composed of wood, flexible polyurethane (PU), polyester (PE), and polypropylene (PP).

@@

Using the same data, the focus of this analysis was to look at trends within the aggregate data. During each test, the fire was ignited and developed until the temperature 1.83~m (6~ft) below the ceiling in the gas cooling room reached or exceeded 250~$^{\circ}$C (482~$^{\circ}$F).  This condition was evidence that a hot gas layer at least 1.83~m deep had filled the room next to the burn room.  Once this criteria was met, gas cooling began. Figure~\ref{fig:gas_cooling_exp4} shows 17 cooling events at one thermocouple array during a continuous burn sequence.  No water or CAF was flowed into the burn room.

@@

This study focused on two key types of fire experiments: gas cooling and fire suppression. In each type of experiment, a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates. In both types of experiments, there were no significant measurable differences between the use of plain water or compressed air foam. To expand upon the 2012 and 2013 tests described by Mitchell~\cite{Mitchell:1}, additional fire suppression experiments were conducted. One set of experiments used only wood-based fuels to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. A final experimental series was conducted in a larger fire room with more ventilation. In this final series, another change was made: the nozzle was hand held and could be moved as needed to effectively extinguish the fire. This section discusses the overall results of the study and provides potential explanations to support them.

@@

While CAF has been readily accepted for use in the wildland and wildland urban interface, the effectiveness or advantages of CAF for interior structural fire fighting was undocumented by experimental data. Questions remain regarding whether CAF provides improved gas cooling when compared to plain water. During the planning of the experiments for this project, there were a range of hypotheses regarding potential outcomes. Some technical panel members thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAF. Their position was based on the fact that CAF has injected air. Because air is a great insulator, it slows heat transfer and results in less cooling. It was their belief that the reduced gas cooling capabilities of CAF may have led to firefighter injuries during interior fire attacks. Other members of the technical panel thought that the bubble structure would provide the same mass flow rate of water but with more surface area. The increased surface area would enable faster heat transfer rate and therefore improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling. However, CAF has not been widely used for structure fires.

@@

Due to inefficiencies in the application of water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the formula based on floor area yields an application density of 3.3~l/min/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~l/min/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). The fire room compartment volumes in these experiments ranged from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). The required flow rate for the fire room in the single-story structure based on the Iowa State University formula was approximately 87~l/min (23~gpm), and the required flow rate for the fire room in the lower level of the two-story structure was approximately 168~l/min (44~gpm). During the experiments, the flow rate was approximately 454~l/min (120~gpm), which exceeds the required flow rate of the Iowa State Formula. The fact that all of the fires were quickly controlled during the experiments is consistent with this methodology.

@@

The fire rooms in these experiments ranged in floor area from approximately 27~m$^2$ (291~ft$^2$) to 51.4~m$^2$ (553~ft$^2$). Applying the NFA formula based on floor area yields a required flow rate for the single-story fire room of 369~l/min (97~gpm). The required flow rate for the fire room in the lower level of the two-story would be approximately 700~l/min (184~gpm). Comparing the hose stream flow rates to the NFA formula requirements indicates that 454~l/min (120~gpm) was sufficient for the smaller room fire in the single-story but was below the required flow rate of approximately 700~l/min (184~gpm) for the large fire compartment in the two-story structure. This is not consistent with the observations of a rapid fire knockdown with both water and CAF hose streams flowing 454~l/min (120~gpm).

Due to recent research conducted by NIST and the UL Firefighter Safety Research Institute (FSRI), the fire service has a renewed interest in controlling the air available to a ventilation limited compartment fire. Limiting the air (oxygen) available for combustion is one means of limiting the heat release rate. If the heat release rate of a fire is reduced due to a lack of oxygen needed for combustion, then in theory, the amount of water needed to cool and suppress the fire should also be reduced.

@@

In all the suppression experiments, the fire was controlled with a total flow of 227~L (60~gal) or less of agent. During the course of the suppression experiments, the fuel packages were modified to incorporate more wood-based fuels. During the planning of the experiments, some of the technical panel members felt that synthetic fuels, such as polyurethane foam, burned mostly at the surface and were therefore easily cooled and suppressed. Wood-based fuels, however, develop a char layer. The burning of wood-based fuels would be ``deep seated'', which might provide a more significant suppression challenge. The reduced surface tension of Class A foam solution could allow the solution to better penetrate and cool the burning wood-based fuels. It was thought that with more exposed burning wood surfaces in the fuel load, CAF would have a better the opportunity to demonstrate a difference in suppression. As shown in Section~\ref{sec:fuel_load}, the wood fuel load consisted of stacks of wood pallets, plywood lined walls, and an exposed wood ``ceiling'' that was constructed of wood ``i-beams'' and OSB. The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

The final set of suppression experiments used a larger space than the previous two sets of suppression experiments in the single-story structure. As discussed previously, with the larger area and double the ventilation of the previous experiments, the fire should have been more challenging to the hose streams. There was also another variable that was be introduced: a firefighter controlling the nozzle. Recall that in the single-story experiments, a monitor nozzle was used in a fixed position. As a result, after the initial fire knockdown, it was typical that the agent was not able to hit all surface of the burning fuel sufficiently and a rekindle would occur. This can be seen for both the water and CAF experiments in the data provided in Appendix~\ref{app:fire_suppression}. In some single-story experiments, the fire was given time to redevelop and additional nozzle flows from the fixed positions were conducted to collect more data and to see if the fires could be extinguished. During the additional flows, no significant differences were seen. In all of the single-story experiments firefighters with a hoseline were needed to extinguish any remaining fire and hot spots. For the last set of experiments, the choice was made to use a firefighter so that the nozzle would be moved in an informed manner in order to make the best use of the hose stream to suppress the fire in the lower level of the two-story structure.

The fire fighting crew was from the MCFRS and every member in the crew had been trained in the use of and had experience with using CAF. The firefighter on the nozzle was the same for all the experiments conducted in the two-story structure. The firefighter used the reach of the hose stream to cool the fire gases just inside the doorway as a means to knockdown the flames that were extending through the open double doors of the basement. Then, the hoseline was advanced and the nozzle was moved rapidly to hit the fuels inside the compartment. The nozzle was shut down when there were no flames visible from the doorway. The flow times were all less than 30 seconds. Even though the fire size had increased, a total flow of 227~L (60~gal) or less of agent was used to cool the hot gas layer and suppress each of the fires. This is consistent with the ventilation analysis above. The nozzle movement clearly had a positive impact on suppressing the fire in the 11~m (36~ft) deep fire compartment.

@@

Full-scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental configurations were designed to be representative of residential fire scenarios. %The water and CAF flows were based on those used by a fire department with 35 CAFS front line engines. This same department also experienced injuries to firefighters while using CAF during interior fire fighting operations in residential structures. 
This study was composed of three types of experiments: spray density, gas cooling, and fire suppression. In each type of experiment, a comparison was conducted between plain water and CAF being discharged from nozzles of similar types with similar mass flow rates of water. The flow rate for each water experiment was 120~gpm. For the CAF experiments, 60~cfm of air and a 0.3~\% flow of Class A foam concentrate were added to the water flow.

@@

There were two different geometries used for the fire suppression experiments. Twelve of the suppression experiments were conducted in a single-story structure with an entry hallway that led to a fire compartment with a floor area of 27~m$^2$ (291~ft$^2$). Within this structure, two different types of fuel loads were used. In one case, most of the fuel consisted of synthetic materials in the upholstered furniture, carpet, and padding. The other fuel load used only wood-based fuels to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. In the experiments conducted in the single-story building, a fixed nozzle was flowed from the hallway for approximately 15~s, and then another fixed nozzle located in the fire compartment was flowed for an additional 15~s.

Seven of the suppression experiments were conducted in a two-story structure.  The fire compartment was the full lower level of the structure, and it had an open floor plan with an area of 51.4~m$^2$ (553~ft$^2$). At one end of the room was an open doorway that was 1.8~m (6~ft) wide and 2~m (6.7~ft) high and led to a stairwell. A manual hose stream application was made from outside the doorway on the lower level. In these experiments, there was a combination of wood-based and synthetic fuels. The fires were allowed to transition through flashover and then the initial fire suppression attack was made. In each case, the initial knockdown of the fire required 60~gal of agent or less.

The results from both the gas cooling and the suppression of residential scale room fires showed the effectiveness of water and CAF hose streams to be similar. Under these test conditions both agents were effective. This study did not address the previously documented attributes of CAF for improved exposure protection and improved prevention of a fire re-kindle relative to plain water.

>>>>

\begin{tabbing}
\hspace{1.5in} \= \\
BDP \> Bi-Directional Probe \\
CAF \> Compressed Air Foam \\
CAFS \> Compressed Air Foam System \\
Cal Poly \> California Polytechnic State University, San Luis Obispo \\
DHS \> Department of Homeland Security \\
FEMA \> Federal Emergency Management Administration \\
HF \> Heat Flux \\
IDLH \> Immediate Danger to Life and Health \\
ESTC \> Delaware County Emergency Services Training Center \\
MCFRS \> Montgomery County (MD) Fire and Rescue Service \\
NIST \> National Institute of Standards and Technology \\
OSB \> Oriented Strand Board \\
TC \> Thermocouple \\
TFT \> Task Force Tips \\
UL FSRI \> UL Firefighter Safety Research Institute \\

@@

Full scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental arrangements and results are presented in this report. The results from both the gas cooling and the suppression of residential scale room fires showed the effectiveness of water and CAF hose streams to be similar. Under the test conditions, both agents were effective in reducing the gas temperatures and suppressing fires in a compartment. This study did not address the previously documented attributes of CAF for improved exposure protection and improved prevention of a fire re-kindle relative to plain water.

@@

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A aqueous solution into an aerated form (a blanket of bubbles) that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they had applied the agent so that they could make a more effective application. In the wildland, water for fire fighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire or CAF can be used for an exterior (indirect) fire suppression attack on a structure.

@@

\begin{table}[!ht]
\centering
\captionof{table}{Tests to Assess Class A Fire Fighting Agents Compared to Water}\label{tab:agent_tests}
\begin{tabular}{lcc}
\toprule[1.5pt]
Test                                    & Similar Performance & Enhanced Performance \\
\midrule
Specific Heat                           &  \checkmark         &                      \\
Fuel Cooling and Penetration            &                     &  \checkmark          \\
Mass Retention                          &                     &  \checkmark          \\
Ignition Inhibition                     &                     &  \checkmark          \\
Drop Size                               &  \checkmark         &                      \\
Contact Angle                           &                     &  \checkmark          \\
Tire Fire Suppression                   &  \checkmark         &                      \\
Wood Crib Suppression                   &  \checkmark         &                      \\
Heptane Fire Suppression                &  \checkmark         &                      \\
Magnesium \& Titanium Fire Suppression  &  \checkmark         &                      \\
Smoke Generation                        &  \checkmark         &                      \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland fire fighting, CAFS is typically applied directly to the fuel from a remote location, typically an environment that is not an immediate danger to life and health (non-IDLH environment). In structural fire fighting, crews typically enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer and or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior fire fighting with CAF are better understood.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires while CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of MCFRS issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order and the experience of the MCFRS played a major role in the development of this study.

@@

The fire suppression series included 20 experiments. Twelve experiments were conducted in the single-story structures (6 in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied and included straight stream, solid stream, and narrow fog stream. Six experiments were conducted in the lower level of the two-story structure using both water and CAF from a straight stream nozzle.  In addition, two CAF experiments were conducted in the two-story burn building with a solid stream. The lower level of the two story structure allowed for a fire compartment that was approximately twice as large as the fire compartment in the single story.

@@

Each structure had a burn room, approximately 6.0~m (19.6~ft) by 4.4~m (14.3~ft) and 2.4~m (8.0~ft) in height and a hallway connecting the burn room to the front of the structure via a small entry foyer. The hallway was 0.95~m (3.1~ft) wide and 5.4~m (17.6~ft) long and the entry foyer was 1.2~m (4.1~ft) by 1.9~m (6.1~ft). The ceiling height in the hallway and entry foyer was 2.4~m (8~ft).  The open door on north side from the exterior to the entry foyer was 2.0~m (6.5~ft) high and 0.9~m (2.9~ft) wide. The opening on the south face of the structure from the exterior to the burn room was 2.4~m (7.75~ft) high and 1.09~m (3.6~ft) wide. A basic plan view of the structure is presented in Figure~\ref{fig:Test_Structure_Floor_Plan}, and a fully dimensioned plan view of the structure is included in in Appendix~\ref{app:floor_plans}.

@@

A second story was added to the west single-story structure in order to conduct suppression experiments in a larger fire compartment than the fire compartment in the single story and to examine the impact of the fire and suppression in a remote compartment (upstairs) that was not directly exposed to the hose stream (see Figure~\ref{fig:delco_2story}).

@@

The fuel load for test numbers 1 through 8 consisted of two sleeper sofas, two cushioned chairs, carpet, padding, and wood paneling along the southeast corner walls. An image of the fuel arrangement is shown in Figure~\ref{fig:Furniture_Fuel_Load}. The sofas and chairs were set up to mimic a typical seating area with one couch against the south wall and one couch against the east wall, each positioned 0.9~m (36~in) away from the southeast corner. A chair was placed adjacent to each couch. The specific locations can be found in the dimensioned schematic, Figure~\ref{fig:Furniture_Fuel_Load_Dimensions}. In total, this fuel package contains approximately 336~kg (739~lb) of fuel consisting of wood, flexible polyurethane (PU), polyester (PE),and polypropylene (PP). Table~\ref{tab:Fire_Suppression_Fuel_Masses} provides a breakdown on the fuel composition for these experiments.

@@

The fuel load for experiments 13 and 14 consisted of three sofas with an average mass of 48.7~kg; 51.4~m$^2$ (553~ft$^2$) of carpet and padding; and approximately 38 sheets of OSB paneling with dimensions 16~mm (0.63~in) x 1.2~m (4~ft) x 2.4~m (8~ft) along the east, west, and south walls and ceiling. The three couches were aligned along the west wall and spaced 1.2~m (4~ft) apart. The specific locations can be found in the dimensioned schematic Figure~\ref{fig:furniture_2story}. In total, this fuel package was composed of wood, flexible polyurethane (PU), polyester (PE), and polypropylene (PP).

@@

Using the same data, the focus of this analysis was to look at trends within the aggregate data. During each test, the fire was ignited and developed until the temperature 1.83~m (6~ft) below the ceiling in the gas cooling room reached or exceeded 250~$^{\circ}$C (482~$^{\circ}$F).  This condition was evidence that a hot gas layer at least 1.83~m (6~ft) deep had filled the room next to the burn room.  Once this criteria was met, gas cooling began. Figure~\ref{fig:gas_cooling_exp4} shows 17 cooling events at one thermocouple array during a continuous burn sequence.  No water or CAF was flowed into the burn room.

@@

This study focused on two key types of fire experiments: gas cooling and fire suppression. In each type of experiment, a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates. In both types of experiments, there were no significant measurable differences between the use of plain water or compressed air foam. To expand upon the 2012 and 2013 tests described by Mitchell~\cite{Mitchell:1}, additional fire suppression experiments were conducted. One set of experiments used only wood-based fuels to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. A final experimental series was conducted in a larger fire room with more ventilation. In this final series, another change was made: the nozzle was hand held and could be moved as needed to completely extinguish the fire. This section discusses the overall results of the study and provides potential explanations to support them.

@@

While CAF has been readily accepted for use in the wildland and wildland urban interface, the effectiveness or advantages of CAF for interior structural fire fighting was undocumented by experimental data. Questions remained regarding whether CAF provides improved gas cooling when compared to plain water. During the planning of the experiments for this project, there were a range of hypotheses regarding potential outcomes. Some technical panel members thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAF. Their position was based on the fact that CAF has injected air. Because air is a great insulator, it might slow the heat transfer rate between the water and the fire gases and result in a reduced cooling rate. It was their belief that the reduced gas cooling capabilities of CAF may have led to firefighter injuries during interior fire attacks. Other members of the technical panel thought that the bubble structure would provide the same mass flow rate of water but with more surface area. The increased surface area would enable faster heat transfer rate and therefore improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling. However, CAF has not been widely used for structure fires.

@@

For the formula as written, it is assumed that the fire compartment is "fully involved" and that the fire is ventilation limited.  Due to inefficiencies in the application of water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the formula based on floor area yields an application density of 3.3~l/min/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~l/min/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). The fire room compartment volumes in these experiments ranged from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). The required flow rate for the fire room in the single-story structure based on the Iowa State University formula was approximately 87~l/min (23~gpm), and the required flow rate for the fire room in the lower level of the two-story structure was approximately 168~l/min (44~gpm). During the experiments, the flow rate was approximately 454~l/min (120~gpm), which exceeds the required flow rate of the Iowa State Formula. The fact that all of the fires were quickly controlled during the experiments is consistent with this methodology.

@@

The fire rooms in these experiments ranged in floor area from approximately 26~m$^2$ (280~ft$^2$) to 51.4~m$^2$ (553~ft$^2$). Applying the NFA formula based on floor area yields a required flow rate for the single-story fire room of 352~l/min (93~gpm). The required flow rate for the fire room in the lower level of the two-story would be approximately 700~l/min (184~gpm). Comparing the hose stream flow rates to the NFA formula requirements indicates that 454~l/min (120~gpm) was sufficient for the smaller room fire in the single-story but was below the required flow rate of approximately 700~l/min (184~gpm) for the large fire compartment in the two-story structure. This is not consistent with the observations of a rapid fire knockdown with both water and CAF hose streams flowing 454~l/min (120~gpm).

\subsection{Ventilation Limited Fires}
Due to recent research conducted by NIST and the UL Firefighter Safety Research Institute (FSRI), the fire service has a renewed interest in controlling the air available to a ventilation limited compartment fire. Limiting the air (oxygen) available for combustion is one means of limiting the heat release rate. If the heat release rate of a fire is reduced due to a lack of oxygen needed for combustion, then in theory, the amount of water needed to cool and suppress the fire should also be reduced.

@@

\subsection{Fuel Loads: Natural vs Synthetic}
In all the suppression experiments, the fire was controlled with no more than 227~L (60~gal) or less of plain water or water in the form of CAF. During the course of the suppression experiments, the fuel packages were modified to incorporate more wood-based fuels. During the planning of the experiments, some of the technical panel members felt that synthetic fuels, such as polyurethane foam, burned mostly at the surface and were therefore easily cooled and suppressed. Wood-based fuels, however, develop a char layer. The burning of wood-based fuels would be ``deep seated'', which might provide a more significant suppression challenge. The reduced surface tension of Class A foam solution could allow the solution to better penetrate and cool the burning wood-based fuels. It was thought that with more exposed burning wood surfaces in the fuel load, CAF would have a better the opportunity to demonstrate a difference in suppression. As shown in Section~\ref{sec:fuel_load}, the wood fuel load consisted of stacks of wood pallets, plywood lined walls, and an exposed wood ``ceiling'' that was constructed of wood ``i-beams'' and OSB. The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

\subsection{Fire Compartment Size}
The final set of suppression experiments used a larger space than the previous two sets of suppression experiments in the single-story structure. As discussed previously, with the larger area and double the ventilation of the previous experiments, the fire should have been more challenging to the hose streams. There was also another variable that was be introduced: a firefighter controlling the nozzle. Recall that in the single-story experiments, a monitor nozzle was used in a fixed position. As a result, after the initial fire knockdown, it was typical that the agent was not able to hit all surface of the burning fuel sufficiently and a rekindle would occur. This can be seen for both the water and CAF experiments in the data provided in Appendix~\ref{app:fire_suppression}. In some single-story experiments, the fire was given time to redevelop and additional nozzle flows from the fixed positions were conducted to collect more data and to see if the fires could be extinguished. During the additional flows, no significant differences were seen. In all of the single-story experiments firefighters with a hoseline were needed to extinguish any remaining fire and hot spots. For the last set of experiments, the choice was made to use a firefighter so that the nozzle would be moved in an informed manner in order to make the best use of the hose stream to suppress the fire in the lower level of the two-story structure.

The fire fighting crew was from the MCFRS and every member in the crew had been trained in the use of CAF and they had experience with using CAF. The firefighter on the nozzle was the same for all the experiments conducted in the two-story structure. The firefighter used the reach of the hose stream to cool the fire gases just inside the doorway as a means to knockdown the flames that were extending through the open double doors of the basement. Then the hoseline was advanced to the doorway and the nozzle was moved rapidly to hit the fuel surfaces inside the compartment. The nozzle was shut down when there were no flames visible from the doorway. The flow times were all less than 30 seconds. Even though the fire size had increased, a total flow of 227~L (60~gal) or less of agent was used to cool the hot gas layer and suppress each of the fires. This is consistent with the ventilation analysis above. The nozzle movement clearly had a positive impact on suppressing the fire in the 11~m (36~ft) deep fire compartment.

@@

Full-scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental configurations were designed to be representative of residential fire scenarios.  The experience of the MCFRS played a major role in the development of this study.  %The water and CAF flows were based on those used by a fire department with 35 CAFS front line engines. This same department also experienced injuries to firefighters while using CAF during interior fire fighting operations in residential structures. 
This study was composed of three types of experiments: spray density, gas cooling, and fire suppression. In each type of experiment, a comparison was conducted between plain water and CAF being discharged from nozzles of similar types with similar mass flow rates of water. The flow rate for each water experiment was 120~gpm. For the CAF experiments, 60~cfm of air and a 0.3~\% flow of Class A foam concentrate were added to the water flow.

@@

There were two different geometries used for the fire suppression experiments. Twelve of the suppression experiments were conducted in single-story structures with an entry hallway that led to a fire compartment with a floor area of 26~m$^2$ (280~ft$^2$). Within these structures, two different types of fuel loads were used. In one case, most of the fuel consisted of synthetic materials in the upholstered furniture, carpet, and padding. The other fuel load used only wood-based fuels to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. In the experiments conducted in the single-story buildings, a fixed nozzle was flowed from the hallway for approximately 15~s, and then another fixed nozzle located in the fire compartment was flowed for an additional 15~s.

Seven of the suppression experiments were conducted in a two-story structure.  The fire compartment was the full lower level of the structure, and it had an open floor plan with an area of 51.4~m$^2$ (553~ft$^2$). At one end of the room was an open doorway that was 1.8~m (6~ft) wide and 2~m (6.7~ft) high and led to a stairwell. A manual hose stream application was made from outside the doorway on the lower level. In these experiments, there was a combination of wood-based and synthetic fuels. The fires were allowed to transition through flashover and then the initial fire suppression attack was made. In each case, the initial knockdown of the fire required no more than 60~gal of of plain water or water in the form of CAF.

The results from both the gas cooling and the suppression of residential scale room fires showed the effectiveness of water and CAF hose streams to be similar. Under the test conditions, both water and CAF were effective in rapidly reducing the gas temperatures and suppressing fires in a compartment.  This study did not address the previously documented attributes of CAF for improved exposure protection and improved prevention of a fire re-kindle relative to plain water.
